---
title: 'Reproducible Research: '
output:
  html_document:
    fig_caption: yes
    keep_md: yes
---
## Peer Assessment #1
### Loading and preprocessing the data

_Conditionally (if not already present) download, unzip, and read-in data._

```{r loading}
library(dplyr, warn.conflicts=FALSE)
library(knitr)
dataURL <-                                  ## Download URL
  paste0("https://d396qusza40orc.cloudfront.",  
         "net/repdata%2Fdata%2Factivity.zip")
targFileName <- paste0("data/repdata\ ",    ## Local data file name
                       "data\ activity.zip")
dataFileName <- "data/activity.csv"         ## Unzipped data file name

if(!file.exists(dataFileName)) {            ## If not present, download file
  download.file(dataURL,                    ## from web and decompress 
               destfile = targFileName, 
               method = "curl")
  unzip(targFileName)
}

if(!"actDat" %in% ls()){                    ## If not already loaded,
  actDat <- read.csv(dataFileName)          ## read-in csv data file.
  actDat$date <- as.Date(actDat$date,       ## Change date to Posix format
                       format="%Y-%m-%d")
}
rm(dataURL, targFileName, dataFileName)     ## Cleanup temp data objects
```


### What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day:  
```{r meanStepsPlot}
plot(                                       ## H-plot the total steps/day
  na.omit(                                  ## Omits NaN and NA values
    summarize(
      group_by(
        actDat, date), 
          sum(steps, na.rm=TRUE))), 
  type="h",
  ylab="Total Steps",
  xlab="Date")
dev.copy(png, file = "meanStepsPlot.png")   ## Save plot to separate file
dev.off()
```

Calculate and report the mean and median total number of steps taken per day:
```{r meanMedSteps}
print(
  paste(
    "Average Steps Per Day",
    mean(na.omit(summarize(group_by(actDat, ## Computes the mean of the sums
                                    date),  ## for each day.  Omits NaN and
                           sum(steps,       ## NA values.
                               na.rm=TRUE)))$sum)))
print(paste("Median Steps Per Day",
            median(na.omit(                 ## Computes the mid-point of the
                     summarize(             ## sums for each day.  Omits NaN 
                       group_by(actDat,     ## and NA values.
                                date), 
                       sum(steps, 
                           na.rm=TRUE)))$sum)))
```   

### What is the average daily activity pattern?  

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis)  

```{r plotStepsInterval}
plot(na.omit(summarize(group_by(            ## L-plot the average steps/interval
                                actDat,     ## Omits NaN and NA values
                                interval), 
                       mean(steps, na.rm=TRUE))), 
                       type="l",
                       ylab="Total Steps",
                       xlab="5-Minute Interval")
dev.copy(png, file =                        ## Save plot to separate file
           "plotStepsInterval.png")
dev.off()
```  

Which 5-minute interval, on average across all the days in the dataset, contains
the maximum number of steps?

```{r maxStepsInterval}
totals <-                                   ## Vector of interval sums
  summarize(
            group_by(actDat, interval), 
            sum(steps, na.rm=TRUE))
maxStep <-                                  ## Largest of vector.
  totals[totals$sum == max(totals$sum), c(1,2)]
hour <-                                     ## Hour component of timestamp for
  as.integer(maxStep$interval/60)           ## largest entry.
minute <-                                   ## Minute component of timestamp for
  ((maxStep$interval/60 - hour)/100) *60    ## vector entry.
print(paste0("Maximum activity is ", 
            maxStep$sum, " steps at ", 
            hour, ":",
            formatC(minute, flag=0,         ## Format minutes.
                    width=5)))
rm(totals, maxStep, hour, minute)           ## Explicit cleanup unused vars.
```
### Input missing values  

Calculate and report the total number of missing values in the dataset (i.e. the 
total number of rows with NAs).  

```{r naValues}
# note that there is no such thing as "imputting"
sum(as.numeric(is.na(actDat[,1])))
```   
Devise a strategy for filling in all of the missing values in the datase. Create 
a new dataset that is equal to the original dataset but with the missing data 
filled in.  

_It is easily determined that NA values are clustered for entire days: in other
words, many days have no observations for steps._   

```{r showNADays}
for(date in unique(actDat$date)){           ## Code chunk demonstrates that
  print(                                    ## values for step observations are
    paste0(                                 ## sometimes NA for full days by not
         mean(actDat[actDat$date==date, 1]),## using mean(na.omit). Computes the
         " ",                               ## step mean for each day of 
         date))                             ## collections. Displays mean with
  }                                         ## the POSIX unformatted date.
```  
_A valid solution would be to discard records with no valid entries for the day.
However, discarding the bad reading dates would not accomplish the instructions
from the assignment:_   

######"Create a new dataset that is equal to the original dataset but with the 
###### missing data filled in."   

_The best compromise solution would be to replace observation step NA values 
with 0 so as not to change the overall average:_  

```{r fillNA}
naReplace <-function(steps){                ## Substitutes 0 for NA values.
  if(is.na(steps)){                         ## Returns a vector of steps.
    return(0)
    }
  else{ 
    return(steps)
    }
  }
newActDat <-                                ## Create a new data frame with the
  data.frame(steps = as.numeric(            ## same sample labels.  Replace all
                      lapply(               ## NA values with the mean of all
                        actDat$steps,       ## days of data collections.
                        naReplace)), 
             date=actDat$date, interval=actDat$interval)
dim(newActDat)                              ## Show the size of the new data
head(newActDat)                             ## Show a sample of the new data
```   

Make a histogram of the total number of steps taken each day and Calculate and 
report the mean and median total number of steps taken per day. Do these values 
differ from the estimates from the first part of the assignment? What is the 
impact of imputing _(sic)_ missing data on the estimates of the total daily 
number of steps?   

_Histogram below:_   

```{r corrDatHist}                          
plot(summarize(group_by(newActDat,          ## H-plot the total steps/day.
                        date),              ## Omits NaN and NA values, same as
               sum(steps)),                 ## above.
  type="h",
  ylab="Total Steps",
  xlab="Date")
dev.copy(png, file = "corrDatHist.png")     ## Save plot to separate file
dev.off()
```  
  
_Mean and Median daily values_  

```{r corrMeanMid}
print(paste("Average Steps Per Day",
            mean(summarize(group_by(        ## Computes and displays the mean of
                             newActDat,     ## the sums for each day.  Omits NaN
                              date),        ## and NA values.
            sum(steps))$sum)))
print(paste("Median Steps Per Day",
            median(summarize(group_by(      ## Computes and displays the mid-
                               newActDat,   ## point of the sums for each day.
                               date),       ## Omits NaN and NA values.
            sum(steps))$sum)))
```   

_Since the NA data were replaced with zero, the daily sums for the affected days 
have not changed.  The most recent histogram shows the same values as that of 
the original data, since the original plot omitted NA values_  

### Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and
“weekend” indicating whether a given date is a weekday or weekend day.  

```{r newFactor}
newActDat <- 
  mutate(newActDat,                         ## Adds new variable for weekday 
         weekday = !grepl("S[a-z]*day",     ## as a boolean. Uses regexp to 
                          as.character(     ## detect day names beginning with
                            lapply(         ## "S" and ending in "day."
                              newActDat$date, 
                              weekdays))))
dim(newActDat)
head(newActDat)
```  

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-
minute interval (x-axis) and the average number of steps taken, averaged across 
all weekday days or weekend days (y-axis).  

```{r panelPlot}
par(mfrow = c(2,1),                         ## Sets up multipanel plots and 
    pin = c(10, 10),                        ## other aesthetics.
    mar = c(.5,2,0,0),
    oma = c(1,1,0,0),
    lab = c(10, 6, 5))
plot(summarize(group_by(newActDat[          ## L-plot the average steps/interval
                          newActDat$weekday==TRUE,],
                        interval),          ## for weekdays. Omits NaN and NA
               mean(steps)),                ## values.         
  type = "l",
  xaxt = "n",                               ## Omits x axis for the top graph
  ylim = c(0,210),
  xlim = c(0,2355),                         ## but sets x axis interval to match
  cex.axis = .7)
text(1150, 210, "Weekday",pos = 1, cex = .8)## Labels plot data
par(mar = c(2,2,0,0))                       ## Modifies the margin for plot 2
plot(summarize(                             ## L-plot the average steps/interval
       group_by(
         newActDat[newActDat$weekday==FALSE,],
         interval),                         ## for weekend data. Omits NaN and
               mean(steps)),                ## NA values.
  type="l",
  ylim = c(0,210),
  xlim = c(0,2355),
  cex.axis = .7 )
text(1150, 210, "Weekend", pos = 1,         ## Labels the plot 
     cex = .8)                     
mtext("Number of Steps",                    ## Labels the shared axes
      side = 2, outer = TRUE)
mtext("Interval", 
      side = 1, outer = TRUE)
dev.copy(png, file = "panelPlot.png")       ## Save plot to separate file
dev.off()
rm(newActDat)                               ## Cleanup temp data frame
```  
________________________________________________________________________________
